Clinical prediction with localized modeling using similarity-based cohorts: A systematic scoping review

### 2025-03-03 docs/Localized-Modeling-BMCAI/localized-modeling-bmcai.Rmd
track bmcai submission & dai revision

---
title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
authors: 
  - name: Adi Cohen
    email: ac4569@mynsu.nova.edu
    affiliation: College of Osteopathic Medicine, Nova Southeastern University
    location: Davie, FL
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: Smathers Libraries Academic Research & Consulting Services, University of Florida
    location: Gainesville, FL
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: Laboratory for Systems Medicine, University of Florida
    location: Gainesville, FL
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) have given rise to a family of techniques we call “localized models”, in which a statistical model is fitted to a neighborhood of labeled cases matched by similarity to a target case. We aim to describe clinical and health applications of localized models to date and propose a general framework for their design and evaluation.

  Methods:
  We searched four bibliographic platforms during 2021 July 19–22, updated 2024 January 24. We set four eligibility criteria to identify applications of localized models to clinical and health tasks. Two authors divided title/abstract screening and reviewed screened entries for inclusion. We discussed settings, tasks, and tools; identified and tabulated themes; and synthesized the methods into a general framework.

  Results:
  Of 1,657 search results, 360 were reviewed, then combined with 43 publications that seeded the review and 1 obtained by citation tracking. 27 were included, published 1997–2022. The specificity of search terms was poor, and inter-rater reliability was low. Almost all models were predictive, the most common tasks being prognosis and diagnosis. Most studies used clinical, occasionally laboratory and image, data. Several addressed memory and runtime costs. A general technique that specializes to most of those reviewed involved matching, retrieval, fitting, and evaluation steps that could optionally be supervised, optimized, or recursively performed.

  Conclusions:
  Localized models have potential to improve the performance of clinical decision support tools while maintaining interpretability, but rigorous comparisons to competing methods must be conducted and computational hurdles must be overcome. We hope that our review will spur future work on efficiency, reproducibility, and user needs.

keywords:
  - case-based reasoning
  - localized modeling
  - nearest neighbors
  - clinical decision support
  - scoping review
bibliography:
  - SR-bibliography.bib
  - SR-reviewers.bib
  - SR-0-Found.bib
  - SR-2-Included.bib
  - SR-3-Tracked.bib
  - SR-5-Screened.bib
biblio-style: unsrt
output:
  rticles::arxiv_article:
    keep_tex: true
    latex_engine: xelatex
    # citation_package: natbib
    citation_package: "default"
header-includes: |
  \usepackage{lscape}
  \newcommand{\landscapebegin}{\begin{landscape}}
  \newcommand{\landscapeend}{\end{landscape}}
---

@@

# Introduction

@@

# Discussion

@@

Because we excluded studies that used conventional NN prediction, many included studies reported new approaches to adaptation subsequent to retrieval. Very few proposed novel similarity measures, possibly because larger studies tended to be reported in separate articles detailing experiments with specific components of the process. However, this also suggests that few experimental studies have focused on the unified development of new retrieval and adaptation strategies.
We also note that the majority of studies evaluated and compared methods using leave-one-out cross-validation (LOOCV). This is an appropriate technique when available data are scarce, and indeed most data sets used by included studies numbered in the hundreds of cases or fewer. This likely follows from the older age of many included studies and from their consistent primary focus on clinical data, which is more costly to collect and comes with more restrictions on its use.

@@

## Conclusions

@@

@@

## Data and code

Search results are publicly available in a Zotero group library.
Data collected or encoded for included studies are publicly available in two Google Sheets.
Code used to analyze these data and generate this manuscript is publicly available in a GitHub repository.

* Search results, included studies, and other bibliography: <https://www.zotero.org/groups/5017571/imsr/>
* Bibliographic and methodological properties of included studies: <https://docs.google.com/spreadsheets/d/1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo/>
* Terminology and composite techniques of included studies: <https://docs.google.com/spreadsheets/d/1xvDJwiLBoI2oz8fxHJ5MjNmiju_RAlK7RJv-wXe1DAs/>
* Code used to conduct analyses and prepare the manuscript: <https://github.com/corybrunson/imsr>

>>>>

---
# Supported options: 
#   sn-nature:       Style for submissions to Nature Portfolio journals
#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
#   sn-mathphys:     Math and Physical Sciences Reference Style
#   sn-aps:          American Physical Society (APS) Reference Style
#   sn-vancouver:    Vancouver Reference Style
#   sn-apa:          APA Reference Style 
#   sn-chicago:      Chicago-based Humanities Reference Style
#   default:         Default

classoptions: 
  - sn-mathphys      
  - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  # - referee       # Optional: Use double line spacing
  # - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour

title: "Clinical prediction with localized modeling using similarity-based cohorts: A systematic scoping review"
titlerunning: "Clinical prediction with localized modeling: A scoping review"

authors: 
  - firstname: Adi
    lastname: Cohen
    degrees: BS
    affiliation: 1

  - firstname: Patti
    lastname: McCall-Junkin
    degrees: MA, MLS
    affiliation: 2

  - firstname: Jason Cory
    lastname: Brunson
    degrees: PhD
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    corresponding: TRUE

affiliations:
  - number: 1
    info:
      orgdiv: College of Osteopathic Medicine
      orgname: Nova Southeastern University
  - number: 2
    info:
      orgdiv: Smathers Libraries Academic Research & Consulting Services
      orgname: University of Florida
  - number: 3
    corresponding: TRUE
    info:
      orgdiv: Laboratory for Systems Medicine
      orgname: University of Florida
    address:
        street: P.O. Box 100225 JHMHC
        city: Gainesville
        state: Florida
        country: USA

keywords:
  - case-based reasoning
  - localized modeling
  - nearest neighbors
  - clinical decision support
  - scoping review

pacs: 
  jel:
    - "C38"
    - "C65"
  msc:
    - "62P10"

# Sample for unstructured abstract 
# abstract: |
#   The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.

# Sample for structured abstract
abstract: |
  **Background:**
  Applications of classical case-based reasoning (CBR) have given rise to a family of techniques we call “localized models”, in which a statistical model is fitted to a neighborhood of labeled cases matched by similarity to a target case. We aim to describe clinical and health applications of localized models to date and propose a general framework for their design and evaluation.

  **Methods:**
  We searched four bibliographic platforms during 2021 July 19–22, updated 2024 January 24. We set four eligibility criteria to identify applications of localized models to clinical and health tasks. Two authors divided title/abstract screening and reviewed screened entries for inclusion. We discussed settings, tasks, and tools; identified and tabulated themes; and synthesized the methods into a general framework.

  **Results:**
  Of 1,657 search results, 360 were reviewed, then combined with 43 publications that seeded the review and 1 obtained by citation tracking. 27 were included, published 1997–2022. The specificity of search terms was poor, and inter-rater reliability was low. Almost all models were predictive, the most common tasks being prognosis and diagnosis. Most studies used clinical, occasionally laboratory and image, data. Several addressed memory and runtime costs. A general technique that specializes to most of those reviewed involved matching, retrieval, fitting, and evaluation steps that could optionally be supervised, optimized, or recursively performed.

  **Conclusions:**
  Localized models have potential to improve the performance of clinical decision support tools while maintaining interpretability, but rigorous comparisons to competing methods must be conducted and computational hurdles must be overcome. We hope that our review will spur future work on efficiency, reproducibility, and user needs.

bibliography:
  - SR-bibliography.bib
  - SR-reviewers.bib
  - SR-0-Found.bib
  - SR-2-Included.bib
  - SR-3-Tracked.bib
  - SR-5-Screened.bib

header-includes: |
  %% Per the spinger doc, new theorem styles can be included using built in style, 
  %% but it seems the don't work so commented below
  %\theoremstyle{thmstyleone}%
  \newtheorem{theorem}{Theorem}%  meant for continuous numbers
  %%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
  %% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  \newtheorem{proposition}[theorem]{Proposition}%
  %%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

  %% \theoremstyle{thmstyletwo}%
  \theoremstyle{remark}
  \newtheorem{example}{Example}%
  \newtheorem{remark}{Remark}%

  %% \theoremstyle{thmstylethree}%
  \theoremstyle{definition}
  \newtheorem{definition}{Definition}%

  \usepackage{lscape}
  \newcommand{\landscapebegin}{\begin{landscape}}
  \newcommand{\landscapeend}{\end{landscape}}

output: rticles::springer_article
---

@@

# Background

@@

# Conclusions

@@

Because we excluded studies that used conventional NN prediction, many included studies reported new approaches to adaptation subsequent to retrieval. Very few proposed novel similarity measures, possibly because larger studies tended to be reported in separate articles detailing experiments with specific components of the process. However, this also suggests that few experimental studies have focused on the unified development of new retrieval and adaptation strategies.
We also note that the majority of studies evaluated and compared methods using LOOCV. This is an appropriate technique when available data are scarce, and indeed most data sets used by included studies numbered in the hundreds of cases or fewer. This likely follows from the older age of many included studies and from their consistent primary focus on clinical data, which is more costly to collect and comes with more restrictions on its use.

@@

## Closing

@@

# List of abbreviations {-}

* CBR: case-based reasoning
* CIS: clinical information system
* CDS: clinical decision support
* AI: artificial intelligence
* NN: nearest neighbors
* GLM: generalized linear model
* ML: machine learning
* DT: decision tree
* PMC: PubMed Central
* LOOCV: leave-one-out cross-validation
* GSA: gravitational search algorithm
* DML: distance metric learning

# Declarations {-}

## Ethics approval and consent to participate {-}

Not applicable.

## Consent for publication {-}

Not applicable.

## Availability of data and materials {-}

Search results, included studies, and other cited sources are publicly available in a Zotero group library:
<https://www.zotero.org/groups/5017571/imsr/>

Data collected or encoded for included studies are publicly available in two Google Sheets:

* Bibliographic and methodological properties of included studies: <https://docs.google.com/spreadsheets/d/1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo/>
* Terminology and composite techniques of included studies: <https://docs.google.com/spreadsheets/d/1xvDJwiLBoI2oz8fxHJ5MjNmiju_RAlK7RJv-wXe1DAs/>

Code used to analyze these data and prepare this manuscript is publicly available in a GitHub repository:
<https://github.com/corybrunson/imsr>

## Competing interests {-}

The authors declare that they have no competing interests.

## Funding {-}

Not applicable.

## Authors' contributions {-}

JCB conceived of the study.
All authors contributed to the study design---search, screen, review, and synthesis.
AC and JCB conducted the search, screen, review, and synthesis (coded and analyzed bibliographic and methodological data).
AC and JCB drafted, and all authors contributed to, read, and approved, the manuscript.

## Acknowledgements {-}

Two anonymous reviewers provided valuable feedback on a previous version of this manuscript.

@@
